The Journey from Wild to Text Book Data: A Case Study of the National Longitudinal Survey of Youth

### 2021-09-01 Report/report_alt.Rmd
Modify the EDA.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      #cache = TRUE, 
                      #cache.path = "cache/",
                      fig.path = "figures/")
library(tidyverse)
library(brolgar)
library(patchwork)
library(kableExtra)
library(MASS)
library(janitor)
library(DiagrammeR)
library(rsvg)
library(webshot)
library(mgcv)
library(modelr)
library(tsibble)
library(lme4)
library(modelr)
```

@@

The same way was also deployed to tidy the rate of wage by year and by ID. The difference is that the hourly rate had only one version of each year. The hours of work and the hourly rate were then joined to calculate the number of jobs that a respondent has and their mean hourly wage. Some observations had 0 in their hourly rate, which is considered as invalid value. Thus, their hourly rate set to be N.A. the result of the joining variables could be seen in the following tibble.

```{r tidy-rate, echo = FALSE}
get_rate <- function(year) {
  new_data_qnames %>%
    dplyr::select(CASEID_1979,
            starts_with("HRP") &
              ends_with(as.character(year))) %>%
    pivot_longer(!CASEID_1979, names_to = "job", values_to = "rate_per_hour") %>%
    separate("job", c("job", "year"), sep = -4) %>%
    mutate(job = paste0("job_0", substr(job, 4, 4))) %>%
    rename(id = CASEID_1979)
}
rates <- list()
for(ayear in c(1979:1994, 1996, 1998, 2000, 2002, 2004, 2006, 2008, 2010, 
               2012, 2014, 2016, 2018)) {
   rates[[ayear]] <- get_rate(ayear)
}
rates_all <- bind_rows(!!!rates)
# join hours and rates variable
hours_wages <- left_join(rates_all, 
                         hours_all, 
                         by = c("id", "year", "job")) %>%
  # set the 0 value in rate_per_hour as NA 
  mutate(rate_per_hour = ifelse(rate_per_hour == 0, NA,
                                rate_per_hour))
head(hours_wages)
```

@@

```{r age-table, echo = FALSE}
age_table <- yowie::demographic_nlsy79 %>%
  group_by(age_1979) %>%
  count(age_1979)

@@

```{r gender-race-table, echo = FALSE}
gender_race_table <- yowie::demographic_nlsy79 %>%
  tabyl(gender, race) %>%
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(gender = ifelse(gender == "MALE", "Male",
                         ifelse(gender == "FEMALE", "Female", "Total")))

@@

```{r wages-tsibble, echo = FALSE}
# create the wages data as tsibble
wages_hs2020_tsibble <- wages_hs2020 %>%
  as_tsibble(key = id,
             index = year,
             regular = FALSE)
```

```{r slope-fig, echo = FALSE, fig.cap="Regression slope of mean hourly wage regression by highest grade completed. The regression model has mean hourly wage as response variable and year as predictor. Respondents who completed 12th grade has the highest median of slope."}

wages_slope <- key_slope(wages_hs2020_tsibble, 
            mean_hourly_wage ~ year) %>%
  left_join(demographic_nlsy79, by = "id") %>%
  mutate(hgc = factor(hgc, levels = c("1ST GRADE", "2ND GRADE", "3RD GRADE", 
                                      "4TH GRADE", "5TH GRADE", "6TH GRADE", 
                                      "7TH GRADE", "8TH GRADE", "9TH GRADE",
                                      "10TH GRADE", "11TH GRADE", "12TH GRADE", 
                                      "UNGRADED"))) %>%
  filter(hgc != "UNGRADED")

ggplot(wages_slope, aes(y = hgc,
                      x = .slope_year)) +
  geom_boxplot()
```

In the second example, we examined the pattern of mean hourly wage corresponding to different groups of gender and race. We took 240 samples of females and males equally and spread them into twelve facets, as shown in Figure \ref{fig:plot-gender} using `facet_strata` function from `brolgar` [@brolgar]. We learn that the mean hourly wage of females and males fluctuated over the years. For example,  in facets 4, 5, and 10, males tend to earn more wages than females.

```{r plot-gender, echo = FALSE, warning = FALSE, fig.cap= "The pattern of mean hourly wages overtime displayed in 12 facets coloured by gender. The respondents belong to high school cohort are randomly sampled. We learn that both male and female respondents had a fluctuate mean hourly wage over the year." }
# sample the data by gender

set.seed(31251587)
wages_hs2020_male <- wages_hs2020_tsibble %>%
  filter(gender == "MALE") %>%
  sample_n_keys(size = 120)

set.seed(31251587)
wages_hs2020_female <- wages_hs2020_tsibble %>%
  filter(gender == "FEMALE") %>%
  sample_n_keys(size = 120)

wages_hs2020_sample_gender <- bind_rows(wages_hs2020_male, wages_hs2020_female)

ggplot(wages_hs2020_sample_gender, 
       aes(x = year,
           y = mean_hourly_wage,
           group = id,
           color = gender)) + 
  geom_line(alpha = 0.7) +
  scale_color_brewer(palette = "Dark2") +
  facet_strata(along = ~gender) 
```

To further investigate this, we calculated the median wage of females relative to males, which is called gender wage gap. According to @OECD, gender pay gap is the difference between median wage of males and females relative to median wage of males. We find that, as shown in \ref{fig:se-gender}, over the years, females tend to earn fewer than males. For example, in 2018 females earn about 78 cents for every $1 earned by males. The widest gender gaps were observed in 1990s.

@@

wages_hs2020_grouped_gender <- wages_hs2020 %>%
  group_by(gender, year) %>%
  summarise(mean = mean(mean_hourly_wage),
            median = median(mean_hourly_wage),
             sd = sd(mean_hourly_wage),
             samplesize = n()) %>%
  ungroup() %>%
  mutate(upper = mean + sd/sqrt(samplesize),
         lower = mean - sd/sqrt(samplesize))

wage_female <- wages_hs2020_grouped_gender %>%
  filter(gender == "FEMALE") %>%
  dplyr::select(year, median) %>%
  rename(female = median)

wage_male <- wages_hs2020_grouped_gender %>%
  filter(gender == "MALE") %>%
  dplyr::select(median) %>%
  rename(male = median)

>>>>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      #cache = TRUE, 
                      #cache.path = "cache/",
                      fig.path = "figures/")
library(tidyverse)
library(brolgar)
library(patchwork)
library(kableExtra)
library(MASS)
library(janitor)
library(DiagrammeR)
library(rsvg)
library(webshot)
library(mgcv)
library(modelr)
library(tsibble)
library(lme4)
library(modelr)
library(forcats)
library(yowie)
```

@@

The same way was also deployed to tidy the rate of wage by year and by ID. The difference is that the hourly rate had only one version of each year. The hours of work and the hourly rate were then joined to calculate the number of jobs that a respondent has and their mean hourly wage. Some observations had 0 in their hourly rate, which is considered as invalid value. Thus, their hourly rate set to be NA. The result of the joining variables could be seen in the following tibble.

```{r tidy-rate, echo = FALSE}
get_rate <- function(year) {
  new_data_qnames %>%
    dplyr::select(CASEID_1979,
            starts_with("HRP") &
              ends_with(as.character(year))) %>%
    pivot_longer(!CASEID_1979, names_to = "job", values_to = "rate_per_hour") %>%
    separate("job", c("job", "year"), sep = -4) %>%
    mutate(job = paste0("job_0", substr(job, 4, 4))) %>%
    rename(id = CASEID_1979)
}
rates <- list()
for(ayear in c(1979:1994, 1996, 1998, 2000, 2002, 2004, 2006, 2008, 2010, 
               2012, 2014, 2016, 2018)) {
   rates[[ayear]] <- get_rate(ayear)
}
rates_all <- bind_rows(!!!rates)
# join hours and rates variable
hours_wages <- left_join(rates_all, 
                         hours_all, 
                         by = c("id", "year", "job")) %>%
  # set the 0 value in rate_per_hour as NA 
  mutate(rate_per_hour = ifelse(rate_per_hour == 0, NA,
                                rate_per_hour),
         hours_work = ifelse(hours_work == 0 | hours_work > 84, NA, hours_work))

head(hours_wages)
```

@@

```{r age-table, echo = FALSE}
age_table <- yowie::demog_nlsy79 %>%
  group_by(age_1979) %>%
  count(age_1979)

@@

```{r gender-race-table, echo = FALSE}
gender_race_table <- yowie::demog_nlsy79 %>%
  tabyl(gender, race) %>%
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(gender = ifelse(gender == "MALE", "Male",
                         ifelse(gender == "FEMALE", "Female", "Total")))

@@

```{r}
wages <- yowie::wages %>%
  mutate(year0 = year - 1979,
         ln_wages = log(wage))

wages_slope <- key_slope(wages, ln_wages ~ year0) %>%
  left_join(yowie::demog_nlsy79, by = "id") %>%
  mutate(hgc_regroup = fct_collapse(hgc, `ungraded` = "UNGRADED",
                                    `8TH grade or less` = c("1ST GRADE", "3RD GRADE",
                                                                 "4TH GRADE", "5TH GRADE",
                                                                 "6TH GRADE", "7TH GRADE",
                                                                 "8TH GRADE"),
                                    `9TH to 11TH grade` = c("9TH GRADE",
                                                            "10TH GRADE",
                                                            "11TH GRADE"),
                                    `12TH grade` = "12TH GRADE")) %>%
  mutate(hgc_regroup = factor(hgc_regroup, levels = c("8TH grade or less", 
                                                      "9TH to 11TH grade", 
                                                      "12TH grade")))

EDA:

Exploring the wages slope of each individual:
  - By sex
  - By education 
  - Exploring the wages slope pattern based on the number of observation

- Slope by sex

```{r}
slope_bottom_5<- wages_slope %>%
  arrange(.slope_year0) %>%
  slice_head(n = 5) %>%
  arrange(desc(.slope_year0)) %>%
  mutate(note = "bottom 5") %>%
  mutate(rank = seq_along(1:5))

slope_top_5<- wages_slope %>%
  arrange(desc(.slope_year0)) %>%
  slice_head(n = 5) %>%
  mutate(note = "top 5") %>%
  mutate(rank = seq_along(1:5))

keep_contrast <- rbind(slope_bottom_5, slope_top_5) %>%
  dplyr::select(.slope_year0, id, note, rank, hgc_regroup)

contrast_slope <- left_join(keep_contrast, wages, by = "id") %>%
  arrange(.slope_year0) %>%
  mutate(id = factor(id, levels = c("7918", "11695", "9176", "10119", "6464", "1271", 
                                    "12371", "9039", "10217", "7782")),
         note = factor(note, levels = c("top 5", "bottom 5")))

ggplot(contrast_slope, aes(x = year, y = ln_wages, color = gender)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(note ~ rank) +
  scale_color_manual(values = c("#04B4AE", "#BF00FF"))

```{r}
ggplot(contrast_slope, aes(x = year, y = ln_wages, color = hgc_regroup)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(note ~ rank)
```

```{r}
wages_obs <- wages %>%
  features(wage, n_obs) %>%
  left_join(wages_slope, by = "id")

ggplot(wages_obs, aes(x = n_obs, y = .slope_year0)) +
  geom_point() +
  geom_jitter()
```

According to @OECD, gender pay gap is the difference between median wage of males and females relative to median wage of males. We find that, as shown in \ref{fig:se-gender}, over the years, females tend to earn fewer than males. For example, in 2018 females earn about 78 cents for every $1 earned by males. The widest gender gaps were observed in 1990s.

@@

wages_gender <- as.tibble(wages) %>%
  group_by(gender, year) %>%
  summarise(mean = mean(wage),
            median = median(wage),
            sd = sd(wage),
            samplesize = n()) %>%
  ungroup() %>%
  mutate(upper = mean + sd/sqrt(samplesize),
         lower = mean - sd/sqrt(samplesize))

wage_female <- wages_gender %>%
  filter(gender == "FEMALE") %>%
  dplyr::select(year, median) %>%
  rename(female = median)

wage_male <- wages_gender %>%
  filter(gender == "MALE") %>%
  dplyr::select(median) %>%
  rename(male = median)
